Speed Test: Data.table Vs. Data.Frame
========================================================


Learn and train to be efficiency in any disipline is a skill to develop which can be acomplish in many ways.  Because surveillance is to constantly monitor health information from a diverse data set in a constant manner process speed is important.

Many years ago I tried and some how managed to analyze millions of rows from health insurance data sets.  By that time I had S-plus and Excel as tools in a laptop with 512 MB of RAM.  The task was such a hassle that at some point a get depressed.  About two years ago I was dealing with simillar data sets.  For the task I needed to managed, reshape and merge, among other one data.frame of 6 million rows 23 columns with other of about two millons rows and seven columns.  I had an HP work station 4 processor 4 GB of RAM.  The task, with many tricks was doable. Nevetheless, I requested other 16 GB of RAM. :)

During the past month, with the programming and R skill that I have aquired throughout the years, I was redefining some management and analysis scripts with the mentioned data sets. Certainly, I was not happy with how the programm perform. Furthermore I started to search and read about performance in R.  Something that I usually do from time to time (I am a huge fan of performance).

Among many new stuff that brilliant **R** developers are doing I decided to implement in my work the [data.table][1.dt] package. The purpose of the developers in doing this package is to extend a data.frame capabilities for fast indexing, fast ordered joins, fast assignment, fast grouping and list columns.

One of the dificulties of using data.table is that it change the paradigm of how manage the data set and how to write programs in **R**. Furthermore, time needs to be invested in order to efficiently use this package.

Bellow is a series of procedures that I have implemented in my script. Those will show you how I used data.table and how the performance improved when compare with traditional data.frame procedures.


[1.dt]: http://cran.r-project.org/web/packages/data.table/index.html


First, load require libraries
```{r libraries, message=FALSE, warning=FALSE}
library(data.table)
library(bit64)
```

To test the speed of the functions I will use system.time() which results are represented in seconds.
## 1
The first test was loading a data.frame using the traditional and useful read.csv vs. the new data.table fread(). As you will notice bellow the 
```{r loadingCsvData, message=FALSE, warning=FALSE}

# data.frame loading .csv data
system.time(
DF <-  read.csv(
  "/media/truecrypt2/ORP2/HIC/ASES/Data/Raw/Claims/reclamaciones_asma_diabetes.csv", 
  header = TRUE, sep = ",")
  )


# data.table loading .csv data
system.time(
DT <-  fread(
    "/media/truecrypt2/ORP2/HIC/ASES/Data/Raw/Claims/reclamaciones_asma_diabetes.csv")
  )



```{r SetNames}
# data.frame
system.time(
  colnames(DF) <- tolower(names(DF))
  )
# data.table
system.time(
setnames(DT, tolower(names(DT)))
  )
```

```{r Colum Selectio}
# data.frame
system.time(
  DF[, 
     c("att_spec", "mpi", "claim_id", "pcp_prov", "icd9_diag_01",
               "risk_type", "primary_carrier_id", "bill_type", "enc_type",
               "proc_code", "rev_code", "municipality", "municipality2", 
               "gender", "dob", "region", "adm_date", "dis_date", "from_date",
               "to_date", "entry_date", "entry_date_key", "date_accident"),]
  )


# data.table
system.time(
  DT[, 
             c("att_spec", "mpi", "claim_id", "pcp_prov", "icd9_diag_01",
               "risk_type", "primary_carrier_id", "bill_type", "enc_type",
               "proc_code", "rev_code", "municipality", "municipality2", 
               "gender", "dob", "region", "adm_date", "dis_date", "from_date",
               "to_date", "entry_date", "entry_date_key", "date_accident"),
with = FALSE]
)

```
```{r Creating Date Colum}
# data.frame
system.time(
  DF$adm_year <- as.Date(DF$adm_date, "%m/%d/%y")
  )

# data.table
system.time(
  DT[, adm_year := as.IDate(DT$adm_date, "%m/%d/%Y")]
  )
```

```{r Perform a function}
system.time(
table(DF$adm_year)
)


system.time(
  DT[, table(adm_year)]
  )

```
```{r}
# data.frame
system.time(
  DF[DF$adm_year == 2010,]
  )

setkey(DT, adm_year)
system.time(
DT[J(2010)]
)
```


```{r Perform_function2}
system.time(
table(DF$adm_year)
)


system.time(
  DT[, table(adm_year)]
  )

```
```{r}
setkey(DT, adm_year)
system.time(
DT[adm_year %in% c(2008, 2009, 2010)]
)
```

```{r substr()}
system.time(
DF$att_spec2 <- substr(DF$att_spec, 1, 3)
)

# data.table
system.time(
DT[, att_spec2 := substr(DT$att_spec, 1, 3)]
)
```


